Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-05-14 latex/prelims.tex
Some minor changes, plus an explanation of how to modify clustered to non clustered
Done?

%% \subsection{Overview}
In this section I will present notations that will be used throughout
this article. These notations are summarized in Section
\ref{subsec:notation}. The theoretical foundations for inverse
problems over function spaces can be found in \cite{Stuart10}.

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectively). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator of the example in Section
\ref{section:intro} is a prime example). Take a Gaussian prior $\param
\sim \pr = \normal(\prmean ,\prcov)$ with some appropriate covariance
operator $\prcov$ on $\hilp$ \cite{Stuart10}. Note that $\fwd \prcov
\fwd^*$ is the prior covariance in $\hilo$ \cite{Stuart10}. As such,
$\fwd \prcov \fwd^*$ is assumed invertible --- an assumption which
will be used below. Let $m$ the number of measurements
taken. Measurements are taken via the \emph{measurement
operator}\footnote{It is common for the measurement and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires that $\fwd$ and $\obs$ are explicitly
separated as in \cite{attia2022stochastic, cvetkovic2023choosing}.}
$\obs \in ( \hilo^* )^m$, whose entries $\meas_j, j=1,\dots,m$ are
called \emph{measurement vector}s:

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error $\eps' \sim
\normal(0,\modcov)$ with $\modcov: \hilo \to \hilo$ a covariance
operator. Observation error is denoted $\eps \sim \normal(0, \sigma^2
I_m)$, with $I_m \in \mathbb{R}^{m \times m}$ the identity. Both error
terms and the prior are assumed independent of each other. Thus, data
is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}
%% \subsubsection{Error Terms}
It is easy to verify that $\obs \eps' + \eps \in \R^m$ is a centered
Gaussian random vector with covariance matrix

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
%% and $e_j \in \R^m$ is the $j$th standard basis vector.
%% The explicit dependence on $\obs$ will be mostly dropped for
%% convenience, so $\Sigma(\obs) = \Sigma$.  Thus, for a fixed $\obs$
%% (i.e.\ a fixed set of observations), $\data = \obs \fwd \param +
%% \bar{\eps}$ with $\bar{\eps} \sim \normal(0,\Sigma)$.
Taking $\modcov = 0$ is a common practice
\cite{Tarantola05,Kaipio2006,Vogel02} and then $\Sigma = \sigma^2I_m$
is a scalar matrix and does not depend on $\obs$.

@@

\subsection{Bayesian D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes expected KL divergence between
posterior and prior measures, $\post$ and $\pr$. It is useful to first
recall the definition of KL divergence:% for an arbitrary prior
measure:
$$
D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
$$

@@

%% , and just like in finite dimensions, a D-optimal design maximizes
%% the Kullback-Leibler divergence between posterior and prior
%% measures For a linear inverse problem with Gaussian prior and
%% Gaussian noise, a D-optimal design minimizes an expression
%% analogous to the determinant of the posterior covariance operator
%% \cite{AlexanderianGloorGhattas14} (also \cref{thm:d optimality}
%% below).  D-optimal designs in infinite-dimensional Hilbert spaces
%% were investigated in \cite{AlexanderianGloorGhattas14}.
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$ are constrained to
  some allowed set of measurement vectors in $\hilo^*$.
\end{definition}

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix\footnote{The frequentist version of
the D-optimality criterion aims to minimize the determinant of the
Fisher information matrix \cite[page 16]{Ucinski05}. Thus, the
Bayesian criterion is just a regularized version of the frequentist
criterion \cite{Chaloner1995}.}. Theorem \ref{thm:d optimality} and
Definition \ref{def:d optimality} have a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}} \\
    &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
Since $\prcov$ is constant, a D-optimal design minimizes the posterior
covariance determinant, analogously to the finite-dimensional case.

\subsection{Notation Summary}\label{subsec:notation}
  Our notation is summarized below. We let:
  \begin{itemize}
  \item $\hilp, \hilo$ Hilbert spaces.
  \item $\fwd:\hilp \to \hilo$ a linear compact operator.
  \item $\pr \sim \mathcal{N}(0, \prcov)$ prior Gaussian measure on
    $\hilp$, with prior covariance operator $\prcov:\hilp \to \hilp$.
  \item $\obs: \hilo \to \mathbb{R}^m$ measurement operator, where $m
    \in \mathbb{N}$ is the number of measurements taken.
  \item $\sigma^2 \in \mathbb{R}_{+}$ observation noise variance.
  \item $\modcov$ model error covariance operator.
    %% $\data = \obs \fwd \param + \eps$, where $\eps \in
    %% \mathbb{R}^m$ isiid $\mathcal{N}(0, \sigma^2)$ noise.
  \item $\Sigma(\obs) = \obs \modcov \obs^* + \sigma^2I$. 
  \item $\post$ the posterior measure, with covariance $\postcov$.
  \item A D-optimality design criterion
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \prcov^{1/2} \fwd ^* \obs^* \Sigma(\obs)^{-1} \obs
        \fwd \prcov^{1/2}).
      \end{split}
    \end{align*}
  %% \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of $\fwd\prcov\fwd^*$
  %%   in decreasing order of magnitude.
  %% %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
  %% \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.
  \end{itemize}

>>>>

%% \subsection{Overview}
In this section we present notations that will be used throughout this
article. These notations are summarized in the definition of Theorem \ref{thm:char}. The theoretical foundations for inverse
problems over function spaces can be found in \cite{Stuart10}.

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces,\footnote{The
subscripts p and o are for ``parameter'' and ``observation'',
respectively.} and denote $\hilo^* \simeq \hilo$ the Hilbert space of
all linear functionals on $\hilo$. Let $\fwd: \hilp \to \hilo$ the
\emph{forward operator}. The forward operator $\fwd$ is assumed linear
and strongly smoothing (has fast decaying modes --- the heat operator
of the 1D heat equation described in Section \ref{section:intro} is a
prime example). Take a Gaussian prior $\param \sim \pr =
\normal(\prmean ,\prcov)$ with some appropriate covariance operator
$\prcov$ on $\hilp$ \cite{Stuart10}. Note that $\fwd \prcov \fwd^*$ is
the prior covariance in $\hilo$ \cite{Stuart10}, and as such, $\fwd
\prcov \fwd^*$ is assumed invertible --- an assumption which will be
used below\footnote{If $\fwd$ has a nontrivial kernel we utilize
Occam's Razor and ignore said kernel.}. Let $m$ the number of
measurements taken. Measurements are taken via the \emph{measurement
operator}\footnote{It is common for the measurement and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires that $\fwd$ and $\obs$ are explicitly
separated as in \cite{attia2022stochastic, cvetkovic2023choosing}.}
$\obs \in ( \hilo^* )^m$, whose entries $\meas_j, j=1,\dots,m$ are
called \emph{measurements}:

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error $\eps' \sim
\normal(0,\modcov)$ with $\modcov$ a covariance operator. Observation
error is denoted $\eps \sim \normal(0, \sigma^2 I_m)$, with $I_m \in
\mathbb{R}^{m \times m}$ the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}
%% \subsubsection{Error Terms}
It is easy to verify that $\obs \eps' + \eps \in \R^m$ is a centered
Gaussian random vector with covariance matrix

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
%% and $e_j \in \R^m$ is the $j$th standard basis vector.
%% The explicit dependence on $\obs$ will be mostly dropped for
%% convenience, so $\Sigma(\obs) = \Sigma$.  Thus, for a fixed $\obs$
%% (i.e.\ a fixed set of observations), $\data = \obs \fwd \param +
%% \bar{\eps}$ with $\bar{\eps} \sim \normal(0,\Sigma)$.
Taking $\modcov = 0$ is a common practice
\cite{Tarantola05,Kaipio2006,Vogel02} and then $\Sigma = \sigma^2I_m$
is a scalar matrix which does not depend on $\obs$.

@@

\subsection{Bayesian D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes expected Kullback-Leibler (KL)
divergence between posterior $\post$ and prior measures $\pr$. It is
useful to first recall the definition of KL divergence \cite{CoverThomas91}:
$$
D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
$$

@@

%% , and just like in finite dimensions, a D-optimal design maximizes
%% the Kullback-Leibler divergence between posterior and prior
%% measures For a linear inverse problem with Gaussian prior and
%% Gaussian noise, a D-optimal design minimizes an expression
%% analogous to the determinant of the posterior covariance operator
%% \cite{AlexanderianGloorGhattas14} (also \cref{thm:d optimality}
%% below).  D-optimal designs in infinite-dimensional Hilbert spaces
%% were investigated in \cite{AlexanderianGloorGhattas14}.
\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
%% It is important to note that since $\obs$ is finite-rank,
%% $\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}$ is
%% trace-class.
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$
  are constrained to some allowed set of measurements in $\hilo^*$.
\end{definition}

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix\footnote{The frequentist version of
the D-optimality criterion aims to minimize the determinant of the
Fisher information matrix \cite[page 16]{Ucinski05}. Thus, the
Bayesian criterion is just a regularized version of the frequentist
criterion \cite{Chaloner1995}.}. Theorem \ref{thm:d optimality} and
Definition \ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of $\prcov$ as constant, so a D-optimal design minimizes the
posterior covariance determinant, analogously to the
finite-dimensional case.

%% \subsection{Notation Summary}\label{subsec:notation}
%%   Our notation is summarized below. We let:
%%   \begin{itemize}
%%   \item $\hilp, \hilo$ Hilbert spaces.
%%   \item $\fwd:\hilp \to \hilo$ a linear compact operator.
%%   \item $\pr \sim \mathcal{N}(0, \prcov)$ prior Gaussian measure on
%%     $\hilp$, with prior covariance operator $\prcov:\hilp \to \hilp$.
%%   \item $\obs: \hilo \to \mathbb{R}^m$ measurement operator, where $m
%%     \in \mathbb{N}$ is the number of measurements taken.
%%   \item $\sigma^2 \in \mathbb{R}_{+}$ observation noise variance.
%%   \item $\modcov$ model error covariance operator.
%%     %% $\data = \obs \fwd \param + \eps$, where $\eps \in
%%     %% \mathbb{R}^m$ isiid $\mathcal{N}(0, \sigma^2)$ noise.
%%   \item $\Sigma(\obs) = \obs \modcov \obs^* + \sigma^2I$. 
%%   \item $\post$ the posterior measure, with covariance $\postcov$.
%%   \item A D-optimality design criterion
%%     \cite{AlexanderianGloorGhattas14}:
%%     \begin{align*}
%%       \begin{split}
%%         \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
%%         % 
%%         % 
%%         % 
%%         &= \frac12 \log \det ( I + \prcov^{1/2} \fwd ^* \obs^* \Sigma(\obs)^{-1} \obs
%%         \fwd \prcov^{1/2}).
%%       \end{split}
%%     \end{align*}
%%   %% \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of $\fwd\prcov\fwd^*$
%%   %%   in decreasing order of magnitude.
%%   %% %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
%%   %% \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.
%%   \end{itemize}
